---
title: "Chart Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chart Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Overview

myIO supports 12 chart types via the `type` argument in `addIoLayer()`. This
vignette shows an example of each.

All examples use built-in R datasets so you can copy and run them directly.

## Line Chart

```{r line, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "line",
    color = "steelblue",
    label = "temperature",
    data = airquality,
    mapping = list(x_var = "Day", y_var = "Temp")
  )
```

## Point (Scatter) Chart

```{r point, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "coral",
    label = "scatter",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  )
```

## Bar Chart

```{r bar, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "bar",
    color = "steelblue",
    label = "bars",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  ) |>
  defineCategoricalAxis(xAxis = TRUE)
```

## Area Chart

```{r area, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "area",
    color = "lightsteelblue",
    label = "area",
    data = airquality,
    mapping = list(x_var = "Day", y_var = "Temp")
  )
```

## Grouped Bar Chart

```{r grouped-bar, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "groupedBar",
    color = "steelblue",
    label = "grouped",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  ) |>
  defineCategoricalAxis(xAxis = TRUE)
```

## Histogram

```{r histogram, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "histogram",
    color = "steelblue",
    label = "hist",
    data = mtcars,
    mapping = list(x_var = "mpg", y_var = "mpg")
  )
```

## Density Plot

```{r density, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "density",
    color = "purple",
    label = "density",
    data = mtcars,
    mapping = list(x_var = "mpg", y_var = "mpg")
  )
```

## Ridgeline Plot

```{r ridgeline, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "ridgeline",
    color = "steelblue",
    label = "ridges",
    data = mtcars,
    mapping = list(x_var = "mpg", y_var = "mpg")
  )
```

## Donut Chart

```{r donut, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "donut",
    color = "steelblue",
    label = "donut",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  )
```

## Gauge Chart

```{r gauge, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "gauge",
    color = "steelblue",
    label = "gauge",
    data = data.frame(value = 75, max = 100),
    mapping = list(x_var = "value", y_var = "max")
  )
```

## Hexbin Chart

```{r hexbin, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "hexbin",
    color = "steelblue",
    label = "hexbin",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  )
```

## Treemap

Treemaps use a different mapping with `level_1` and `level_2` instead of
`x_var` and `y_var`:

```{r treemap, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "treemap",
    label = "cars",
    data = mtcars,
    mapping = list(level_1 = "vs", level_2 = "cyl")
  )
```

## Combining Chart Types

You can overlay multiple chart types on a single widget:

```{r combined, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "steelblue",
    label = "points",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  addIoLayer(
    type = "line",
    color = "orange",
    label = "line",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  addIoStatLayer(
    type = "lm",
    color = "red",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  )
```
