---
title: "Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Overview

This gallery showcases what you can build with myIO. Each example is
self-contained and uses built-in R datasets.

## Scatter Plot with Trend Line

Overlay raw data with a linear model fit:

```{r scatter-trend, }
myIO() |>
  addIoLayer(
    type = "point",
    color = "#E69F00",
    label = "observations",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  addIoLayer(
    type = "line",
    transform = "lm",
    color = "#D55E00",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  setAxisFormat(xLabel = "Weight (1000 lbs)", yLabel = "Miles per Gallon") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Multi-Series Line Chart

Track temperature across months using grouped data:

```{r multi-line, }
aq <- airquality
aq$Month <- paste0("M", aq$Month)

myIO() |>
  addIoLayer(
    type = "line",
    color = c("#0072B2", "#E69F00", "#009E73", "#D55E00", "#CC79A7"),
    label = "Month",
    data = aq,
    mapping = list(x_var = "Day", y_var = "Temp", group = "Month")
  ) |>
  setAxisFormat(xLabel = "Day of Month", yLabel = "Temperature (F)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Categorical Bar Chart

Aggregate data across categories with custom formatting:

```{r bar-chart, }
myIO() |>
  addIoLayer(
    type = "bar",
    color = "#56B4E9",
    label = "avg_mpg",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  ) |>
  defineCategoricalAxis(xAxis = TRUE) |>
  setAxisFormat(
    xAxis = ".0f", yAxis = ".1f",
    xLabel = "Cylinders", yLabel = "Miles per Gallon"
  ) |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Area Chart with Confidence Band

Visualize a range using low and high y values:

```{r area-band, }
aq <- airquality[airquality$Month == 7 & complete.cases(airquality), ]
aq$TempLow <- aq$Temp - 8
aq$TempHigh <- aq$Temp + 8

myIO() |>
  addIoLayer(
    type = "area",
    color = "#56B4E9",
    label = "range",
    data = aq,
    mapping = list(x_var = "Day", low_y = "TempLow", high_y = "TempHigh")
  ) |>
  addIoLayer(
    type = "line",
    color = "#0072B2",
    label = "actual",
    data = aq,
    mapping = list(x_var = "Day", y_var = "Temp")
  ) |>
  setAxisFormat(xLabel = "Day of Month (July)", yLabel = "Temperature (F)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Dark Mode Dashboard

Combine theming with custom colors for a dark aesthetic:

```{r dark-mode, }
myIO() |>
  addIoLayer(
    type = "point",
    color = "#48dbfb",
    label = "scatter",
    data = mtcars,
    mapping = list(x_var = "hp", y_var = "qsec")
  ) |>
  addIoLayer(
    type = "line",
    transform = "lm",
    color = "#ff6b6b",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "hp", y_var = "qsec")
  ) |>
  setTheme(
    text_color = "#b0b0b0",
    grid_color = "#2d2d2d",
    bg = "#0d1117",
    font = "Inter, system-ui, sans-serif"
  ) |>
  setAxisFormat(xLabel = "Horsepower", yLabel = "Quarter Mile Time (s)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Donut Chart

Summarize proportions with a donut:

```{r donut, }
donut_data <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)

myIO() |>
  addIoLayer(
    type = "donut",
    color = c("#E69F00", "#56B4E9", "#009E73"),
    label = "cylinders",
    data = donut_data,
    mapping = list(x_var = "cyl", y_var = "mpg")
  )
```

## Hexbin Density Plot

Visualize point density in large datasets:

```{r hexbin, }
myIO() |>
  addIoLayer(
    type = "hexbin",
    color = "#0072B2",
    label = "density",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg", radius = 20)
  ) |>
  setAxisFormat(xLabel = "Weight (1000 lbs)", yLabel = "Miles per Gallon") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Histogram with Reference Line

Show a distribution with a mean reference line:

```{r histogram-ref, }
myIO() |>
  addIoLayer(
    type = "histogram",
    color = "#CC79A7",
    label = "distribution",
    data = mtcars,
    mapping = list(value = "mpg")
  ) |>
  setReferenceLines(xRef = mean(mtcars$mpg))
```

## Interactive Scatter with Draggable Points

Enable drag interactions for exploratory analysis:

```{r drag, }
myIO() |>
  addIoLayer(
    type = "point",
    color = "#009E73",
    label = "draggable",
    data = mtcars[1:10, ],
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  dragPoints() |>
  setAxisFormat(xLabel = "Weight", yLabel = "MPG") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Treemap

Explore hierarchical data:

```{r treemap, }
myIO() |>
  addIoLayer(
    type = "treemap",
    label = "cars",
    data = mtcars,
    mapping = list(level_1 = "vs", level_2 = "cyl")
  )
```

## Heatmap

Visualize a matrix of values with continuous color:

```{r heatmap, }
df <- expand.grid(
  quarter = c("Q1", "Q2", "Q3", "Q4"),
  segment = c("Enterprise", "Mid-Market", "SMB"),
  stringsAsFactors = FALSE
)
df$revenue <- c(120, 145, 160, 155, 80, 95, 110, 105, 40, 55, 65, 60)

myIO() |>
  addIoLayer(
    type = "heatmap",
    color = "#4E79A7",
    label = "Revenue",
    data = df,
    mapping = list(x_var = "quarter", y_var = "segment", value = "revenue")
  ) |>
  defineCategoricalAxis(xAxis = TRUE, yAxis = TRUE) |>
  setAxisFormat(xLabel = "Quarter", yLabel = "Segment")
```

## Candlestick Chart

Show open-high-low-close financial data:

```{r candlestick, }
df <- data.frame(
  day   = 1:8,
  open  = c(44, 45, 43, 46, 45, 47, 46, 48),
  high  = c(47, 48, 46, 49, 48, 50, 49, 52),
  low   = c(43, 44, 42, 45, 44, 46, 45, 47),
  close = c(45, 43, 46, 45, 47, 46, 48, 51)
)

myIO() |>
  addIoLayer(
    type = "candlestick",
    label = "Stock",
    data = df,
    mapping = list(
      x_var = "day", open = "open",
      high = "high", low = "low", close = "close"
    )
  ) |>
  setAxisFormat(xAxis = ".0f", yAxis = "$,.0f",
                xLabel = "Trading Day", yLabel = "Price") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Waterfall Chart

Show how values accumulate through sequential steps:

```{r waterfall, }
df <- data.frame(
  step     = c("Revenue", "COGS", "Gross Profit", "OpEx", "Tax", "Net Income"),
  value    = c(500, -200, NA, -120, -30, NA),
  is_total = c(FALSE, FALSE, TRUE, FALSE, FALSE, TRUE)
)

myIO() |>
  addIoLayer(
    type = "waterfall",
    label = "P&L Bridge",
    data = df,
    mapping = list(x_var = "step", y_var = "value", total = "is_total")
  ) |>
  defineCategoricalAxis(xAxis = TRUE) |>
  setAxisFormat(yAxis = "$,.0f", xLabel = "Step", yLabel = "Amount") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Sankey Flow Diagram

Visualize flows between nodes:

```{r sankey, }
df <- data.frame(
  source = c("Organic", "Organic", "Paid", "Paid", "Referral"),
  target = c("Landing", "Blog", "Landing", "Promo", "Landing"),
  value  = c(300, 120, 200, 80, 150)
)

myIO() |>
  addIoLayer(
    type = "sankey",
    color = c("#4E79A7", "#F28E2B", "#E15759", "#76B7B2", "#59A14F"),
    label = "Traffic Flow",
    data = df,
    mapping = list(source = "source", target = "target", value = "value")
  )
```

## Boxplot

Compare distributions across groups:

```{r boxplot, }
myIO() |>
  addIoLayer(
    type = "boxplot",
    color = "#4E79A7",
    label = "Petal Width",
    data = iris,
    mapping = list(x_var = "Species", y_var = "Petal.Width")
  ) |>
  setAxisFormat(xLabel = "Species", yLabel = "Petal Width (cm)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Violin Plot

Show distribution shape alongside summary statistics:

```{r violin, }
myIO() |>
  addIoLayer(
    type = "violin",
    color = "#59A14F",
    label = "Sepal Width",
    data = iris,
    mapping = list(x_var = "Species", y_var = "Sepal.Width"),
    options = list(showBox = TRUE, showMedian = TRUE)
  ) |>
  setAxisFormat(xLabel = "Species", yLabel = "Sepal Width (cm)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Ridgeline Plot

Compare distributions across groups with overlapping density curves:

```{r ridgeline, }
df <- mtcars
df$cyl <- as.character(df$cyl)

myIO() |>
  addIoLayer(
    type = "ridgeline",
    color = c("#4E79A7", "#F28E2B", "#E15759"),
    label = "MPG Distribution",
    data = df,
    mapping = list(x_var = "hp", y_var = "mpg", group = "cyl"),
    options = list(overlap = 0.5)
  ) |>
  setAxisFormat(xLabel = "Horsepower", yLabel = "Density") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```
