---
title: "Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Overview

This gallery showcases what you can build with myIO. Each example is
self-contained and uses built-in R datasets.

## Scatter Plot with Trend Line

Overlay raw data with a linear model fit:

```{r scatter-trend, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "#E69F00",
    label = "observations",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  addIoLayer(
    type = "line",
    transform = "lm",
    color = "#D55E00",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  setAxisFormat(xLabel = "Weight (1000 lbs)", yLabel = "Miles per Gallon") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Multi-Series Line Chart

Track temperature across months using grouped data:

```{r multi-line, eval = FALSE}
aq <- airquality
aq$Month <- paste0("M", aq$Month)

myIO() |>
  addIoLayer(
    type = "line",
    color = c("#0072B2", "#E69F00", "#009E73", "#D55E00", "#CC79A7"),
    label = "Month",
    data = aq,
    mapping = list(x_var = "Day", y_var = "Temp", group = "Month")
  ) |>
  setAxisFormat(xLabel = "Day of Month", yLabel = "Temperature (F)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Categorical Bar Chart

Aggregate data across categories with custom formatting:

```{r bar-chart, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "bar",
    color = "#56B4E9",
    label = "avg_mpg",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  ) |>
  defineCategoricalAxis(xAxis = TRUE) |>
  setAxisFormat(
    xAxis = ".0f", yAxis = ".1f",
    xLabel = "Cylinders", yLabel = "Miles per Gallon"
  ) |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Area Chart with Confidence Band

Visualize a range using low and high y values:

```{r area-band, eval = FALSE}
aq <- airquality[complete.cases(airquality), ]
aq$TempLow <- aq$Temp - 8
aq$TempHigh <- aq$Temp + 8

myIO() |>
  addIoLayer(
    type = "area",
    color = "#56B4E9",
    label = "range",
    data = aq,
    mapping = list(x_var = "Day", low_y = "TempLow", high_y = "TempHigh")
  ) |>
  addIoLayer(
    type = "line",
    color = "#0072B2",
    label = "actual",
    data = aq,
    mapping = list(x_var = "Day", y_var = "Temp")
  ) |>
  setAxisFormat(xLabel = "Day", yLabel = "Temperature (F)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Dark Mode Dashboard

Combine theming with custom colors for a dark aesthetic:

```{r dark-mode, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "#48dbfb",
    label = "scatter",
    data = mtcars,
    mapping = list(x_var = "hp", y_var = "qsec")
  ) |>
  addIoLayer(
    type = "line",
    transform = "lm",
    color = "#ff6b6b",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "hp", y_var = "qsec")
  ) |>
  setTheme(
    text_color = "#b0b0b0",
    grid_color = "#2d2d2d",
    bg = "#0d1117",
    font = "Inter, system-ui, sans-serif"
  ) |>
  setAxisFormat(xLabel = "Horsepower", yLabel = "Quarter Mile Time (s)") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Donut Chart

Summarize proportions with a donut:

```{r donut, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "donut",
    color = c("#E69F00", "#56B4E9", "#009E73"),
    label = "cylinders",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  )
```

## Hexbin Density Plot

Visualize point density in large datasets:

```{r hexbin, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "hexbin",
    color = "#0072B2",
    label = "density",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  setAxisFormat(xLabel = "Weight (1000 lbs)", yLabel = "Miles per Gallon") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Histogram with Reference Line

Show a distribution with a mean reference line:

```{r histogram-ref, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "histogram",
    color = "#CC79A7",
    label = "distribution",
    data = mtcars,
    mapping = list(value = "mpg")
  ) |>
  setReferenceLines(x = mean(mtcars$mpg))
```

## Interactive Scatter with Draggable Points

Enable drag interactions for exploratory analysis:

```{r drag, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "#009E73",
    label = "draggable",
    data = mtcars[1:10, ],
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  dragPoints() |>
  setAxisFormat(xLabel = "Weight", yLabel = "MPG") |>
  setMargin(top = 20, bottom = 70, left = 60, right = 10)
```

## Treemap

Explore hierarchical data:

```{r treemap, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "treemap",
    label = "cars",
    data = mtcars,
    mapping = list(level_1 = "vs", level_2 = "cyl")
  )
```
