---
title: "Getting Started with myIO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with myIO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Overview

myIO is an R package that combines R and D3.js to create interactive data
visualizations using the `htmlwidgets` framework. Charts built with myIO work
in RStudio, Shiny apps, and R Markdown documents.

The core workflow is simple:

1. Create a widget with `myIO()`
2. Add layers with `addIoLayer()` or `addIoStatLayer()`
3. Customize with option functions

## Your First Chart

Create a scatter plot of `mtcars` data:

```{r basic-scatter, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "steelblue",
    label = "mpg_vs_wt",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  )
```

## Adding Multiple Layers

You can stack layers on a single chart. Here we add points and a linear
model trend line:

```{r multi-layer, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "steelblue",
    label = "points",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  addIoStatLayer(
    type = "lm",
    color = "red",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  )
```

## Customizing the Chart

Option functions let you control axes, margins, legends, and more. They
all follow the same pipe-friendly pattern:

```{r customize, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "bar",
    color = "orange",
    label = "bars",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  ) |>
  defineCategoricalAxis(xAxis = TRUE) |>
  setAxisFormat(xAxis = ".0f", yAxis = ".0f") |>
  setAxisFormat(xLabel = "Cylinders", yLabel = "Miles per Gallon") |>
  setMargin(top = 40, bottom = 70, left = 60, right = 10) |>
  setTransitionSpeed(speed = 500)
```

## Grouped Data

When your data has a grouping variable, pass it in the `mapping` to
automatically create one layer per group:

```{r grouped, eval = FALSE}
aq <- airquality
aq$Month <- paste0("M", aq$Month)

myIO() |>
  addIoLayer(
    type = "line",
    color = c("steelblue", "orange", "green", "red", "purple"),
    label = "Month",
    data = aq,
    mapping = list(x_var = "Day", y_var = "Temp", group = "Month")
  )
```

## Using in Shiny

myIO provides standard Shiny bindings:
- `myIOOutput()` for the UI
- `renderMyIO()` for the server

```{r shiny, eval = FALSE}
library(shiny)

ui <- fluidPage(
  myIOOutput("my_chart", width = "100%", height = "400px")
)

server <- function(input, output) {
  output$my_chart <- renderMyIO({
    myIO() |>
      addIoLayer(
        type = "point",
        color = "steelblue",
        label = "scatter",
        data = mtcars,
        mapping = list(x_var = "wt", y_var = "mpg")
      )
  })
}

shinyApp(ui, server)
```

## Available Option Functions

| Function | Purpose |
|----------|---------|
| `setAxisFormat()` | Set d3.js axis and tooltip formats, axis labels |
| `setAxisLimits()` | Set min/max for x and y axes |
| `setMargin()` | Set chart margins (top, bottom, left, right) |
| `setColorScheme()` | Set color palette and category labels |
| `setTransitionSpeed()` | Set animation speed (0 to disable) |
| `setReferenceLines()` | Add x and y reference lines |
| `setToggle()` | Add a toggle button to switch y variable |
| `setToolTipOptions()` | Configure tooltip behavior |
| `defineCategoricalAxis()` | Treat axes as categorical |
| `flipAxis()` | Swap x and y axes |
| `dragPoints()` | Make points draggable |
| `suppressAxis()` | Hide x and/or y axes |
| `suppressLegend()` | Hide the legend |
