---
title: "LLM Tool Calling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLM Tool Calling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(myIO)
```

myIO ships a generated schema that agents can use to produce contract-correct chart specs. The schema is generated from the R and JavaScript contracts and is available to both R helpers and a standalone MCP server.

## R helpers

List chart types:

```{r}
head(myio_list_chart_types())
```

Validate a chart spec:

```{r}
spec <- list(
  type = "boxplot",
  mapping = list(column_var = "Species", value_var = "Sepal.Width")
)
myio_validate_spec(spec)
```

Pass a column type map to catch data-contract issues:

```{r}
myio_validate_spec(
  list(type = "point", mapping = list(x_var = "wt", y_var = "mpg")),
  columns = list(wt = "numeric", mpg = "character")
)
```

Validate function-call argument names:

```{r}
myio_validate_call("setAxisFormat", list(axis_x = ".0f"))
myio_function_signature("setAxisFormat")
```

## MCP server

The MCP package is intentionally separate from the R package so CRAN does not need Node dependencies.

```{sh, eval=FALSE}
cd mcp
npm install
node server.mjs
```

Claude Desktop / Claude Code style configuration:

```json
{
  "mcpServers": {
    "myio": {
      "command": "node",
      "args": ["/absolute/path/to/myIO/mcp/server.mjs"]
    }
  }
}
```

Cursor uses the same command and args in its MCP configuration.

The server exposes six tools:

- `list_chart_types`
- `get_chart_schema`
- `validate_spec`
- `list_functions`
- `get_function_signature`
- `validate_call`

## Ollama loop

For local tool-calling loops, expose the same six tools with JSON inputs. A minimal repair flow is:

1. Ask the model for a chart spec.
2. Call `validate_spec`.
3. If `valid` is `FALSE`, feed the error codes and suggestions back to the model.
4. Repeat until the tool returns `valid = TRUE`.
5. Validate any proposed R call with `validate_call`.

Example repair:

```json
{
  "type": "boxplot",
  "mapping": {
    "column_var": "Species",
    "value_var": "Sepal.Width"
  }
}
```

`validate_spec` returns `MISSING_MAPPING` for `x_var` and `y_var`, plus `UNKNOWN_MAPPING_KEY` suggestions for the hallucinated keys. The repaired spec is:

```json
{
  "type": "boxplot",
  "mapping": {
    "x_var": "Species",
    "y_var": "Sepal.Width"
  }
}
```

Validation guarantees contract correctness: known chart type, allowed transform, expected mapping keys, optional data-column compatibility, and known function arguments. It does not guarantee the semantic or aesthetic quality of the selected chart.
