---
title: "LLM Tool Calling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLM Tool Calling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(myIO)
# Render tool results as the JSON an agent actually receives.
as_json <- function(x) {
  cat(jsonlite::toJSON(x, pretty = TRUE, auto_unbox = TRUE, null = "null"))
}
```

myIO is **agent-buildable**. It ships a machine-readable schema generated from
its own R and JavaScript contracts — every chart type, its required mappings and
valid transforms, and every exported function signature. Because the schema is
generated from the engine (never hand-authored), it cannot drift from what the
library actually renders.

Agents read that schema through six tools — available both as R functions
(shown here) and as a standalone MCP server — to produce **contract-correct**
chart specs and to **self-repair** when they get one wrong.

## The full chart surface

The schema covers every chart type the engine can render — not a curated subset:

```{r}
myio_list_chart_types()
```

```{r}
length(myio_list_chart_types())   # chart types
length(myio_list_functions())     # exported function signatures
```

Each type carries its own contract: which mappings are required, which fields
must be numeric, and which transforms are valid for it.

```{r}
as_json(myio_chart_schema("boxplot"))
```

## Validating a chart spec

A model's first attempt frequently hallucinates mapping keys. `myio_validate_spec()`
catches the structural mistakes and returns stable error `code`s plus
fuzzy-matched `suggestion`s:

```{r}
bad_spec <- list(
  type = "boxplot",
  mapping = list(column_var = "Species", value_var = "Sepal.Width")
)
as_json(myio_validate_spec(bad_spec))
```

Feeding those suggestions back (`column_var` → `x_var`, `value_var` → `y_var`)
produces a spec that validates:

```{r}
good_spec <- list(
  type = "boxplot",
  mapping = list(x_var = "Species", y_var = "Sepal.Width")
)
as_json(myio_validate_spec(good_spec))
```

Pass a column type map to also check the spec against real data — mapped columns
must exist, and numeric fields must map to numeric columns:

```{r}
as_json(myio_validate_spec(
  list(type = "point", mapping = list(x_var = "wt", y_var = "mpg")),
  columns = list(wt = "numeric", mpg = "character")
))
```

## Validating function calls

The same applies to `set*()` calls — wrong argument names are a common LLM
mistake. `myio_validate_call()` checks the function name and its arguments
against the real signature:

```{r}
as_json(myio_validate_call("setAxisFormat", list(axis_x = ".0f")))
```

```{r}
myio_function_signature("setAxisFormat")
```

## The six tools

| Tool | R function | Purpose |
|------|------------|---------|
| `list_chart_types` | `myio_list_chart_types()` | Every chart type in the schema |
| `get_chart_schema` | `myio_chart_schema(type)` | One type's contract (or all) |
| `validate_spec` | `myio_validate_spec(spec, columns)` | Validate a chart spec, with optional data-contract checks |
| `list_functions` | `myio_list_functions()` | Every exported function name |
| `get_function_signature` | `myio_function_signature(fn)` | A function's argument names |
| `validate_call` | `myio_validate_call(fn, args)` | Validate a function name + arguments |

All six return the same stable error codes: `UNKNOWN_TYPE`, `MISSING_MAPPING`,
`UNKNOWN_MAPPING_KEY`, `INVALID_TRANSFORM`, `MISSING_COLUMN`,
`NON_NUMERIC_COLUMN`, `UNKNOWN_FUNCTION`, `UNKNOWN_ARGUMENT`.

## A repair loop

The pattern an agent follows, whether it calls the R functions or the MCP tools:

1. Propose a chart spec.
2. Call `validate_spec`.
3. If `valid` is `false`, apply the `suggestion` for each error and go to 2.
4. Once valid, validate any accompanying `set*()` call with `validate_call`.

```{r}
spec <- list(type = "boxplot",
             mapping = list(column_var = "Species", value_var = "Sepal.Width"))
res  <- myio_validate_spec(spec)

while (!res$valid) {
  for (err in res$errors) {
    if (!is.null(err$suggestion) && identical(err$code, "UNKNOWN_MAPPING_KEY")) {
      spec$mapping[[err$suggestion]] <- spec$mapping[[err$field]]
      spec$mapping[[err$field]] <- NULL
    }
  }
  res <- myio_validate_spec(spec)
}
as_json(spec)
```

## MCP server

For agents that speak the Model Context Protocol (Claude Desktop, Claude Code,
Cursor), the same six tools are exposed by a standalone Node server. It is kept
separate from the R package so CRAN never needs Node dependencies.

```{sh, eval=FALSE}
cd mcp
npm install
node server.mjs
```

Client configuration (Claude Desktop / Claude Code / Cursor use the same shape):

```json
{
  "mcpServers": {
    "myio": {
      "command": "node",
      "args": ["/absolute/path/to/myIO/mcp/server.mjs"]
    }
  }
}
```

The MCP server reads the same generated schema, so its results are identical to
the R functions above — enforced by a shared conformance corpus that both
implementations must pass.

## What validation guarantees — and what it does not

`validate_spec` and `validate_call` guarantee **contract correctness**: a known
chart type, an allowed transform, the expected mapping keys, optional
data-column compatibility, and known function arguments. They do **not**
guarantee that the chosen chart is the *right* chart for the question, or that a
mapped column is semantically meaningful — choosing a boxplot when a histogram
was wanted is a valid spec. Semantic and aesthetic judgement remains the model's
job; the tools close the structural-error class.
```