---
title: "Sparklines in Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparklines in Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Overview

myIO's sparkline mode renders compact, axes-free charts at 20px height, ideal
for embedding inline in table cells. Enable it with `sparkline = TRUE`:

```{r basic-sparkline, eval = FALSE}
myIO(data.frame(x = 1:20, y = cumsum(rnorm(20))), sparkline = TRUE) |>
  addIoLayer("line", label = "trend", mapping = list(x_var = "x", y_var = "y"))
```

## Supported types

```{r supported-types, eval = FALSE}
df <- data.frame(x = 1:12, y = c(4, 6, 5, 8, 7, 9, 6, 10, 8, 11, 9, 12))

# Line sparkline
myIO(df, sparkline = TRUE) |>
  addIoLayer("line", label = "line", mapping = list(x_var = "x", y_var = "y"))

# Bar sparkline
myIO(df, sparkline = TRUE) |>
  addIoLayer("bar", label = "bars", mapping = list(x_var = "x", y_var = "y"))

# Area sparkline
myIO(df, sparkline = TRUE) |>
  addIoLayer("area", label = "area", mapping = list(x_var = "x", y_var = "y"))
```

## What sparkline mode does

- Strips axes, legend, reference lines, and all interactions
- Hides the floating action button, so no export/legend panel overlays the mark
- Sets margins to 1px
- Defaults height to 20px, width to 100%
- Only supports `line`, `bar`, and `area` types
