---
title: "Transforms & Theming"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transforms & Theming}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Transforms

Transforms derive new data from an existing layer before rendering. You
specify a transform with the `transform` argument in `addIoLayer()`.

### Available Transforms

| Transform | Description |
|-----------|-------------|
| `"identity"` | Default. Passes data through unchanged. |
| `"lm"` | Fits a linear model and returns fitted values. |

### Linear Model Trend Line

Add a trend line to a scatter plot by combining a `"point"` layer with a
`"line"` layer that uses `transform = "lm"`:

```{r lm-example, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "#E69F00",
    label = "points",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  addIoLayer(
    type = "line",
    transform = "lm",
    color = "red",
    label = "trend",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  )
```

The `lm` transform fits `y ~ x` using `stats::lm()` and replaces the y
values with the fitted values. The resulting line is sorted by x so it
renders as a smooth trend.

### Transform + Type Compatibility

Not every transform works with every chart type. The table below shows
valid combinations:

| Type | Supported Transforms |
|------|---------------------|
| `"line"` | `"identity"`, `"lm"` |
| `"point"` | `"identity"` |
| `"bar"` | `"identity"` |
| `"area"` | `"identity"` |
| `"groupedBar"` | `"identity"` |
| `"histogram"` | `"identity"` |
| `"hexbin"` | `"identity"` |
| `"treemap"` | `"identity"` |
| `"donut"` | `"identity"` |
| `"gauge"` | `"identity"` |

If you pass an incompatible combination, `addIoLayer()` will error with a
clear message.

## Theming

Use `setTheme()` to customize chart appearance with CSS custom properties.

### Basic Theming

```{r theme-basic, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point",
    color = "#56B4E9",
    label = "scatter",
    data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  setTheme(
    text_color = "#e0e0e0",
    grid_color = "#333333",
    bg = "#1a1a2e",
    font = "Fira Code, monospace"
  )
```

### Theme Parameters

| Parameter | CSS Property | Description |
|-----------|-------------|-------------|
| `text_color` | `--chart-text-color` | Axis labels and tick text |
| `grid_color` | `--chart-grid-color` | Grid and axis lines |
| `bg` | `--chart-bg` | Chart background color |
| `font` | `--chart-font` | Font family |

### Custom Properties

Pass additional CSS custom properties via `...` (the `chart-` prefix is
added automatically):

```{r theme-custom, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "bar",
    color = "coral",
    label = "bars",
    data = mtcars,
    mapping = list(x_var = "cyl", y_var = "mpg")
  ) |>
  defineCategoricalAxis(xAxis = TRUE) |>
  setTheme(
    bg = "#fafafa",
    text_color = "#333",
    "border-radius" = "8px"
  )
```

### Dark Mode Example

Combine theming with color choices for a dark-mode chart:

```{r dark-mode, eval = FALSE}
aq <- airquality
aq$Month <- paste0("M", aq$Month)

myIO() |>
  addIoLayer(
    type = "line",
    color = c("#00d2ff", "#ff6b6b", "#feca57", "#48dbfb", "#ff9ff3"),
    label = "Month",
    data = aq,
    mapping = list(x_var = "Day", y_var = "Temp", group = "Month")
  ) |>
  setTheme(
    text_color = "#b0b0b0",
    grid_color = "#2d2d2d",
    bg = "#0d1117",
    font = "Inter, system-ui, sans-serif"
  )
```
