---
title: "Why myIO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why myIO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## The problem

R has excellent interactive visualization packages. So why build another one?

Three things changed:

1. **Plotly is deprioritizing R.** In late 2025, Plotly retired its R documentation and Dash docs for all non-Python languages. The plotly.R package has 750+ open issues, and core features like confidence interval rendering have been broken since 2019 with no maintainer response.

2. **The ggplotly() translation layer is fundamentally limited.** Converting ggplot2 objects to plotly.js traces requires assumptions that break real-world statistical visualizations — CI ribbons render incorrectly, regression equations disappear, and label geoms are silently dropped.

3. **No existing package computes and renders statistical overlays natively.** Users must pre-compute confidence intervals, regression diagnostics, and error bars in R, then manually layer them onto interactive charts. This is tedious and error-prone.

myIO solves this with a **composable transform pipeline**: statistical computation happens in R (where it belongs), rendering happens in D3.js (where it's beautiful), and the two connect through a clean data contract.

## What myIO does differently

### 1. Confidence intervals that actually work

In plotly.R, `geom_smooth(se = TRUE)` converted through `ggplotly()` produces
visual artifacts — nonsensical extra ribbons, broken NA handling, and misaligned
bands. These bugs have been open for 7+ years ([#1472](https://github.com/plotly/plotly.R/issues/1472),
[#1060](https://github.com/plotly/plotly.R/issues/1060)).

myIO computes CI bands natively in R and renders them as a first-class area layer:

```{r ci-example, eval = FALSE}
myIO(data = mtcars) |>
  addIoLayer(type = "point", color = "#4E79A7", label = "Data",
    mapping = list(x_var = "wt", y_var = "mpg")) |>
  addIoLayer(type = "line", color = "#E15759", label = "Trend",
    transform = "lm",
    mapping = list(x_var = "wt", y_var = "mpg")) |>
  addIoLayer(type = "area", color = "#E15759", label = "95% CI",
    transform = "ci",
    mapping = list(x_var = "wt", y_var = "mpg"),
    options = list(level = 0.95))
```

The `ci` transform uses `stats::predict()` with `interval = "confidence"` and
outputs a grid of `low_y` / `high_y` bounds. The area renderer displays them
as a smooth band. Prediction intervals work the same way with
`options = list(interval = "prediction")`.

### 2. One-line regression with R² annotation

In plotly.R, displaying a regression equation requires the `ggpmisc` package's
`stat_poly_eq` — which silently disappears when converted through `ggplotly()`
([#1687](https://github.com/plotly/plotly.R/issues/1687)).

myIO's `regression` composite handles the entire workflow in one call:

```{r regression-example, eval = FALSE}
myIO(data = mtcars) |>
  addIoLayer(type = "regression", label = "MPG vs Weight",
    mapping = list(x_var = "wt", y_var = "mpg"),
    options = list(method = "lm", showCI = TRUE, showStats = TRUE))
```

This auto-expands into four sub-layers: scatter points, trend line, CI band,
and an R² annotation — all styled consistently. Change `method` to `"loess"`
or `"polynomial"` to switch the model.

### 3. Composable transforms

Every myIO chart uses the same pattern: **type + transform = layer**. Transforms
are R functions that compute statistical summaries and return a data frame. The
renderer doesn't know or care what the numbers mean — it just plots fields.

This means you can compose transforms freely:

```{r compose-example, eval = FALSE}
# Scatter + LOESS smooth + CI band + residual plot
myIO(data = mtcars) |>
  addIoLayer(type = "point", label = "Data",
    mapping = list(x_var = "wt", y_var = "mpg")) |>
  addIoLayer(type = "line", label = "LOESS", transform = "loess",
    mapping = list(x_var = "wt", y_var = "mpg"),
    options = list(span = 0.5))
```

Available transforms:

| Transform | What it computes | Output |
|-----------|-----------------|--------|
| `"lm"` | Linear regression fitted values | Line |
| `"loess"` | LOESS non-linear smoothing | Line |
| `"polynomial"` | Polynomial regression (degree N) | Line |
| `"ci"` | Confidence or prediction interval | Area band |
| `"smooth"` | Simple or exponential moving average | Line |
| `"mean"` | Group mean | Points or bars |
| `"mean_ci"` | Group mean ± confidence interval | Range bars (error bars) |
| `"residuals"` | Regression residuals vs. fitted | Points |
| `"summary"` | Aggregation (count, sum, sd, var, min, max) | Points or bars |

All transforms use base R's `stats` package — no additional dependencies.

### 4. Composite charts expand into primitives

Complex statistical charts like boxplots and violins are defined as
**composites** that auto-expand into simpler layers:

```
boxplot → rangeBar (IQR) + point (whisker caps) + point (median) + point (outliers)
violin  → area (density) + rangeBar (IQR box) + point (median)
regression → point (scatter) + line (trend) + area (CI band) + text (R²)
```

Each sub-layer renders independently, which means:

- The IQR box can have a different color than the whiskers
- Outliers can be toggled on/off via `options = list(showOutliers = TRUE)`
- Tooltips work on every sub-component
- Animations are layered (whiskers grow from box edges, then caps appear)

You don't need to know about this decomposition — just call
`addIoLayer(type = "boxplot", ...)` and it works. But if you need custom
control, you can compose the sub-layers manually.

### 5. Error bars with one line

Mean ± confidence interval bars are a standard in scientific publications.
In most interactive viz packages, you must compute the CI yourself and
manually assemble the layers.

myIO does it in one call:

```{r mean-ci-example, eval = FALSE}
myIO(data = iris) |>
  addIoLayer(type = "rangeBar", label = "Mean ± 95% CI",
    transform = "mean_ci",
    mapping = list(x_var = "Species", y_var = "Sepal.Length"),
    options = list(level = 0.95)) |>
  defineCategoricalAxis(xAxis = TRUE) |>
  setAxisFormat(xLabel = "Species", yLabel = "Sepal Length")
```

The `mean_ci` transform computes the group mean, standard error, and
t-distribution CI bounds — matching `stats::t.test()` output exactly.

## Design philosophy

myIO follows three principles:

1. **R computes, D3 renders.** Statistical computation stays in R where the
   ecosystem is strongest. JavaScript handles rendering and interaction.
   There is no translation layer that can break.

2. **Type + transform = layer.** Every chart is built from the same two
   concepts. This makes the API small and predictable — 15 transforms and
   18 chart types compose into hundreds of visualizations.

3. **Composites expand into primitives.** Complex charts decompose into
   simpler ones. This means new statistical overlays don't require new
   renderers — they reuse existing ones.

## When to use myIO

myIO is a good fit when you need:

- Interactive statistical charts in Shiny apps or R Markdown
- Confidence intervals, regression lines, or error bars that render correctly
- Composable overlays (scatter + trend + CI band + annotation)
- Moving average or smoothing overlays on time-series data
- Regression diagnostics (residual plots)
- Publication-quality D3.js rendering with minimal code

## When to use something else

- **ggplot2 + ggiraph**: If you need the full ggplot2 grammar and just want
  hover/click interactivity added on top. ggiraph faithfully preserves ggplot2
  output.
- **echarts4r**: If you need 30+ chart types and don't mind learning a
  non-ggplot2 API. echarts4r has broader chart coverage.
- **plotly**: If you have existing ggplotly() code that works and don't need
  CI bands or statistical annotations.
