---
title: "Linking myIO with Crosstalk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking myIO with Crosstalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = requireNamespace("crosstalk", quietly = TRUE)
)
```

[Crosstalk](https://rstudio.github.io/crosstalk/) is a standard for
widget-to-widget communication that lets users interact with several
visualizations at once — brushing points in one chart dims matching rows
in a companion table, filtering a table hides the non-matching points on
the chart, and so on. myIO participates in this protocol via
`setLinked()`.

## Which myIO layers participate

As of v1.2, nine chart-layer types respond to crosstalk selection and
filter events:

| Layer type | Selection (target) | Filter (target) | Brush source |
|------------|:------------------:|:---------------:|:------------:|
| `point`     | yes | yes | yes (with `setBrush()`) |
| `bar`       | yes | yes | yes (with `setBrush()`) |
| `groupedBar`| yes | yes | yes (with `setBrush()`) |
| `histogram` | yes | yes | yes (with `setBrush()`) |
| `hexbin`    | yes | yes | yes (with `setBrush()`) |
| `waffle`    | yes | yes | — |
| `beeswarm`  | yes | yes | yes (with `setBrush()`) |
| `lollipop`  | yes | yes | — |
| `dumbbell`  | yes | yes | — |

Aggregate layer types (`boxplot`, `violin`, `qq`, `regression`,
`density`, `ridgeline`, `survfit`, `comparison`) do not participate in
v1.2. To drive selection over aggregate views, layer a linked scatter
next to the aggregate chart.

## Selection vs filter

- **Selection** dims non-matching elements (sets their CSS opacity to
  the `--chart-brush-dim-opacity` variable). Other widgets are not
  affected by their layout.
- **Filter** hides non-matching elements entirely (`display: none`).
  Use it when linked partners have a filter UI — e.g. a
  `crosstalk::filter_select()` dropdown — that genuinely narrows the
  dataset.

Pass `filter = TRUE` to `setLinked()` to enable filter-driven hiding.

## Source and target modes

`setLinked(mode = ...)` accepts three values:

- `"source"` — this chart emits selections (via its brush) but does not
  react to incoming ones.
- `"target"` — this chart reacts to incoming selections but does not
  emit its own.
- `"both"` (default) — bidirectional.

## Example 1: scatter + DT

```{r dt-example, eval = requireNamespace("DT", quietly = TRUE) && requireNamespace("crosstalk", quietly = TRUE)}
library(crosstalk)
library(myIO)
library(DT)

shared <- SharedData$new(mtcars, key = ~rownames(mtcars))

bscols(
  widths = c(7, 5),
  myIO(data = shared$data()) |>
    addIoLayer(
      type = "point", label = "cars",
      mapping = list(x_var = "wt", y_var = "mpg")
    ) |>
    setBrush(direction = "xy") |>
    setLinked(shared),
  datatable(shared, rownames = FALSE, options = list(pageLength = 5))
)
```

Brush a rectangle in the chart: the matching rows in the table stay
bold and the rest dim. Click rows in the table: the matching points
stay solid, the rest dim.

## Example 2: lollipop + reactable

```{r reactable-example, eval = requireNamespace("reactable", quietly = TRUE) && requireNamespace("crosstalk", quietly = TRUE)}
library(crosstalk)
library(myIO)
library(reactable)

# Aggregate mtcars by cyl for a simple lollipop
by_cyl <- data.frame(
  cyl = c("4", "6", "8"),
  avg_mpg = c(
    mean(mtcars$mpg[mtcars$cyl == 4]),
    mean(mtcars$mpg[mtcars$cyl == 6]),
    mean(mtcars$mpg[mtcars$cyl == 8])
  )
)
shared <- SharedData$new(by_cyl, key = ~cyl)

bscols(
  widths = c(6, 6),
  myIO(data = shared$data()) |>
    addIoLayer(
      type = "lollipop", label = "by-cyl",
      mapping = list(x_var = "cyl", y_var = "avg_mpg")
    ) |>
    setLinked(shared),
  reactable(shared, selection = "multiple", onClick = "select")
)
```

## Example 3: plotly bridge (static listing)

plotly's own crosstalk integration means you can link a myIO chart
directly to a plotly chart over the same `SharedData`:

```{r plotly-listing, eval = FALSE}
library(crosstalk)
library(myIO)
library(plotly)

shared <- SharedData$new(mtcars, key = ~rownames(mtcars))
bscols(
  myIO(data = shared$data()) |>
    addIoLayer(type = "point", label = "m",
               mapping = list(x_var = "wt", y_var = "mpg")) |>
    setBrush(direction = "xy") |>
    setLinked(shared),
  plot_ly(shared, x = ~hp, y = ~mpg, type = "scatter", mode = "markers")
)
```

Not evaluated in this vignette. Install `plotly` and run locally.

## Example 4: leaflet map (static listing)

```{r leaflet-listing, eval = FALSE}
library(crosstalk)
library(myIO)
library(leaflet)

quakes_small <- head(quakes, 50)
shared <- SharedData$new(quakes_small)
bscols(
  leaflet(shared) |> addTiles() |> addMarkers(),
  myIO(data = shared$data()) |>
    addIoLayer(type = "point", label = "mag",
               mapping = list(x_var = "depth", y_var = "mag")) |>
    setLinked(shared)
)
```

Not evaluated in this vignette. Install `leaflet` and run locally.

## Troubleshooting

- **My boxplot/violin doesn't respond.** Aggregate layer types are not
  linkable in v1.2. Drive selection through a sibling scatter or bar
  layer.
- **My chart is in the linked group but brushing doesn't emit.** Add
  `setBrush()` to the chart layer. Without a brush, the chart can only
  be a target, not a source.
- **Selections don't visually change anything on the linked widget.**
  Confirm the `SharedData$new()` key vector matches the row identity
  the other widget expects. Mismatched keys silently fail.
