---
title: "Interactive I/O in Shiny"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)

embed_app <- function(height = "600px") {
  url <- "https://www.morton-analytics.com/myio/"
  htmltools::tags$iframe(
    src = url,
    width = "100%",
    height = height,
    style = "border: 1px solid #dee2e6; border-radius: 8px;",
    allow = "fullscreen",
    loading = "lazy"
  )
}
```

myIO is not just a charting library — it is a bidirectional data interface.
User interactions (brush, click, annotate, slider) flow back to R as
structured data. This article walks through each pattern with a live demo
and the code behind it.

**Try each interaction in the live app below**, then read the code that
powers it.

## Live Demo App

The full gallery app runs all 20 chart types plus the four I/O interaction
demos. Navigate to **Interactions** in the top menu to try each one.

```{r live-app, echo = FALSE}
embed_app("700px")
```

---

## Brush Selection

Drag a rectangle on the chart to select data points. The selected rows
are returned as a Shiny reactive input — no JavaScript required.

### Try it

Open **Interactions > Brush Selection** in the app above. Drag on the
scatter plot, then watch the sidebar update with the selected count and
data range. Use the direction dropdown to switch between 2D, X-only, and
Y-only brushing.

### Code

```{r brush-code, eval = FALSE}
# UI
fluidRow(
  column(8, myIOOutput("brushPlot", height = "450px")),
  column(4,
    h4("Selected Points"),
    verbatimTextOutput("brushInfo"),
    selectInput("brush_dir", "Brush Direction",
      choices = c("Both axes" = "xy", "X only" = "x", "Y only" = "y"))
  )
)

# Server
output$brushPlot <- renderMyIO({
  myIO() |>
    addIoLayer(type = "point", color = "#4E79A7", label = "Cars",
      data = mtcars, mapping = list(x_var = "wt", y_var = "mpg")) |>
    setBrush(direction = input$brush_dir) |>
    setAxisFormat(xLabel = "Weight (1000 lbs)", yLabel = "Miles per Gallon")
})

output$brushInfo <- renderPrint({
  brushed <- input$`myIO-brushPlot-brushed`
  if (is.null(brushed)) return("Drag on the chart to select points.")
  sel <- jsonlite::fromJSON(brushed)
  if (length(sel$keys) == 0) return("No points selected.")
  cat(length(sel$keys), "of", nrow(mtcars), "points selected\n")
  if (!is.null(sel$extent$x)) {
    cat("X range:", round(sel$extent$x[1], 2), "-", round(sel$extent$x[2], 2), "\n")
  }
  if (!is.null(sel$extent$y)) {
    cat("Y range:", round(sel$extent$y[1], 2), "-", round(sel$extent$y[2], 2), "\n")
  }
})
```

### How it works

1. `setBrush()` adds a D3 brush overlay to the chart
2. On brush end, the selected data points are emitted as a `"brushed"` event
3. In Shiny, the event arrives as `input$myIO-{id}-brushed` — a JSON payload with:
   - `data` — the selected row objects
   - `keys` — `_source_key` values for linking
   - `extent` — brush bounds in data coordinates
4. In static HTML (no Shiny), a status bar shows the selection count and a
   [Clear] button. The CSV export scopes to selected points when
   `on_select = "export"`.

### Supported chart types

Brush works on element-based types: `point`, `bar`, `histogram`, `hexbin`,
`groupedBar`. Line and area charts are excluded — they don't have discrete
selectable elements.

---

## Click-to-Annotate

Click any data point to attach a label. Annotations are stored as
structured data — not cosmetic SVG — and can be exported as a CSV or
read as a Shiny reactive.

### Try it

Open **Interactions > Click-to-Annotate** in the app above. Click any
point in the iris scatter plot. Choose a label from the dropdown and
optionally pick a category color. The annotation appears as a ring +
label on the chart, and the table in the sidebar updates.

### Code

```{r annotate-code, eval = FALSE}
# UI
fluidRow(
  column(8, myIOOutput("annotatePlot", height = "450px")),
  column(4,
    h4("Annotations"),
    tableOutput("annotationTable")
  )
)

# Server
output$annotatePlot <- renderMyIO({
  myIO() |>
    addIoLayer(type = "point", color = "#4E79A7", label = "Iris",
      data = iris, mapping = list(x_var = "Sepal.Length", y_var = "Petal.Length")) |>
    setAnnotation(
      labels = c("outlier", "cluster edge", "typical"),
      colors = c(outlier = "#E63946", `cluster edge` = "#F4A261", typical = "#2A9D8F")
    ) |>
    setAxisFormat(xLabel = "Sepal Length", yLabel = "Petal Length")
})

output$annotationTable <- renderTable({
  ann <- input$`myIO-annotatePlot-annotated`
  if (is.null(ann)) return(data.frame())
  parsed <- jsonlite::fromJSON(ann)
  if (length(parsed$annotations) == 0) return(data.frame())
  df <- parsed$annotations
  data.frame(Label = df$label, X = round(as.numeric(df$x), 2), Y = round(as.numeric(df$y), 2))
})
```

### Annotation data structure

Each annotation is a structured object with these fields:

| Field | Type | Description |
|-------|------|-------------|
| `_source_key` | string | Links back to the original data row |
| `x`, `y` | number | Data coordinates (not pixels) |
| `x_var`, `y_var` | string | Column names for context |
| `label` | string | User-provided label (max 30 chars) |
| `category` | string | CSS color from category picker (or null) |
| `layerLabel` | string | Which chart layer the point belongs to |
| `timestamp` | string | ISO 8601 when the annotation was created |

In static HTML, annotations are exportable via the bottom-sheet CSV button
or the status bar [Export] button.

---

## Linked Brushing

Connect two or more myIO charts via Crosstalk. Brush points in one chart
and watch them highlight in the other — with shared `_source_key` values
linking the rows across views.

### Try it

Open **Interactions > Linked Brushing** in the app above. The left chart
shows `wt vs mpg`, the right shows `hp vs mpg`. Drag a brush on the left
chart — the same cars light up on the right.

### Code

```{r linked-code, eval = FALSE}
library(crosstalk)

# Server
shared <- SharedData$new(mtcars, key = ~rownames(mtcars))

output$linkedA <- renderMyIO({
  myIO() |>
    addIoLayer(type = "point", color = "#4E79A7", label = "wt vs mpg",
      data = shared$data(), mapping = list(x_var = "wt", y_var = "mpg")) |>
    setBrush() |>
    setLinked(shared, mode = "source") |>
    setAxisFormat(xLabel = "Weight", yLabel = "MPG")
})

output$linkedB <- renderMyIO({
  myIO() |>
    addIoLayer(type = "point", color = "#E15759", label = "hp vs mpg",
      data = shared$data(), mapping = list(x_var = "hp", y_var = "mpg")) |>
    setLinked(shared, mode = "target") |>
    setAxisFormat(xLabel = "Horsepower", yLabel = "MPG")
})
```

### How linking works

1. Both charts share the same `crosstalk::SharedData` object
2. The source chart calls `setBrush()` + `setLinked(shared, mode = "source")`
3. The target chart calls `setLinked(shared, mode = "target")`
4. When the source brushes, it sends selected `_source_key` values via
   Crosstalk's `SelectionHandle`
5. The target receives the keys and dims non-matching elements

Use `mode = "both"` for bidirectional linking. Set `filter = TRUE` to
hide non-matching points instead of dimming them.

Crosstalk is a Suggests dependency — it's only loaded when `setLinked()`
is called. Your core myIO code stays dependency-free.

### Linked brushing + recomputation

For deeper analysis, combine Crosstalk (instant visual feedback) with
Shiny reactivity (recomputed statistics):

```{r linked-recompute, eval = FALSE}
# When chart A brushes, also trigger an R recomputation in chart B
observeEvent(input$`myIO-chartA-brushed`, {
  sel <- jsonlite::fromJSON(input$`myIO-chartA-brushed`)
  subset <- mtcars[rownames(mtcars) %in% sel$keys, ]
  output$chartB <- renderMyIO({
    myIO(data = subset) |>
      addIoLayer(type = "regression", label = "refitted",
        mapping = list(x_var = "wt", y_var = "mpg"))
  })
})
```

This gives two-speed feedback: instant highlight via Crosstalk, then a
smooth transition to a refitted model via Shiny.

---

## Parameter Sliders

Add sliders below the chart that control transform parameters. Moving a
slider triggers Shiny re-rendering — the chart recomputes and animates
to the new state.

### Try it

Open **Interactions > Parameter Slider** in the app above. Drag the
confidence level slider and watch the CI band narrow or widen in
real-time as the regression recomputes.

### Code

```{r slider-code, eval = FALSE}
output$sliderPlot <- renderMyIO({
  ci <- input$`myIO-sliderPlot-slider-ci_level`
  if (is.null(ci)) ci <- 0.95

  myIO(data = df) |>
    addIoLayer(type = "regression", label = "Yield Model",
      mapping = list(x_var = "day", y_var = "yield"),
      options = list(method = "lm", showCI = TRUE, level = ci, showStats = TRUE)) |>
    setSlider("ci_level", "Confidence Level", 0.80, 0.99, ci, 0.01) |>
    setAxisFormat(xLabel = "Day of Experiment", yLabel = "Yield (mg)")
})
```

### How it works

1. `setSlider()` renders an HTML range input below the chart
2. On drag, the slider sends its value to `input$myIO-{id}-slider-{param}`
3. This invalidates the `renderMyIO` reactive, which recomputes the chart
   with the new parameter value
4. D3 transitions animate the change smoothly

The slider is **Shiny-only**. In static HTML it renders disabled with a
tooltip explaining the limitation.

### Multiple sliders

`setSlider()` is additive — call it multiple times:

```{r multi-slider, eval = FALSE}
myIO(data = df) |>
  addIoLayer(type = "regression", label = "fit",
    mapping = list(x_var = "x", y_var = "y"),
    options = list(method = "polynomial", degree = deg, level = ci)) |>
  setSlider("ci_level", "Confidence Level", 0.80, 0.99, 0.95, 0.01) |>
  setSlider("degree", "Polynomial Degree", 1, 5, 2, 1)
```

### Debounce

For heavy transforms, increase the debounce to avoid excessive
re-renders:

```{r debounce, eval = FALSE}
setSlider("span", "LOESS Span", 0.1, 1.0, 0.5, 0.05, debounce = 500)
```

---

## Shiny Input Reference

All myIO Shiny inputs follow the pattern `myIO-{outputId}-{event}`:

| Input key | Trigger | Payload |
|-----------|---------|---------|
| `myIO-{id}-brushed` | Brush end or clear | `{ data, extent, keys, layerLabel }` |
| `myIO-{id}-annotated` | Annotation add/remove/clear | `{ annotations, action, latest }` |
| `myIO-{id}-slider-{param}` | Slider drag (debounced) | Numeric value |
| `myIO-{id}-rollover` | Hover on element | JSON data point |
| `myIO-{id}-dragEnd` | Point drag completed | `{ point, layerLabel }` |
| `myIO-{id}-error` | Render error | Error message string |

---

## Static HTML vs Shiny

Not every feature requires a running Shiny server:

| Feature | Static HTML | Shiny |
|---------|-------------|-------|
| Brush selection | Visual highlight + scoped CSV export | + reactive `input$brushed` |
| Annotations | Click to label + CSV export | + reactive `input$annotated` |
| Linked brushing | Crosstalk client-side linking | + server-side recomputation |
| Parameter sliders | Disabled (renders with default) | Full reactive recomputation |
| Drag points | Live regression refit | + reactive `input$dragEnd` |
| Tooltips | Always work | + reactive `input$rollover` |
| CSV/PNG export | Always work | Always work |

The I/O system is designed so that static HTML gets the best experience
possible — brushing and annotation work without Shiny. The Shiny layer
adds reactive data flow on top.
