---
title: "Theme Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theme Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(myIO)
```

## Overview

myIO ships named theme presets that recolor the entire chart — background,
gridlines, text, tooltip, brush, and reference-line colors — in one call:

```{r usage, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point", label = "obs", data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  setTheme(preset = "midnight")
```

Each preview below is the same scatter-plus-trend chart rendered under a
different preset, so you can compare them directly. `"light"` and `"dark"`
are also available via `setTheme(mode = "light" | "dark")`.

```{r presets-list}
myio_theme_presets <- c(
  "light", "dark", "midnight", "ocean", "forest", "sunset",
  "monochrome", "neon", "corporate", "academic", "nature",
  "minimal", "retro", "warm"
)
myio_theme_presets
```

## Live previews

```{r gallery, echo = FALSE}
themed_demo <- function(preset) {
  myIO() |>
    addIoLayer(
      type = "point", label = "observations", data = mtcars,
      mapping = list(x_var = "wt", y_var = "mpg")
    ) |>
    addIoLayer(
      type = "line", transform = "lm", label = "trend", data = mtcars,
      mapping = list(x_var = "wt", y_var = "mpg")
    ) |>
    setTheme(preset = preset) |>
    setTitle(paste0('setTheme(preset = "', preset, '")')) |>
    setAxisFormat(xLabel = "Weight (1000 lbs)", yLabel = "Miles per Gallon") |>
    setMargin(top = 40, bottom = 60, left = 60, right = 10)
}

htmltools::tagList(
  lapply(myio_theme_presets, function(p) {
    htmltools::div(
      style = "margin-bottom: 1.5rem;",
      themed_demo(p)
    )
  })
)
```

## Custom overrides

Presets are a starting point. Layer your own CSS custom properties on top with
`setTheme(overrides = ...)` or the `--`-prefixed `...` arguments:

```{r overrides, eval = FALSE}
myIO() |>
  addIoLayer(
    type = "point", label = "obs", data = mtcars,
    mapping = list(x_var = "wt", y_var = "mpg")
  ) |>
  setTheme(
    preset = "midnight",
    overrides = list("--chart-annotation-ring" = "#22d3ee")
  )
```
