---
title: "Troubleshooting"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Common Errors and Fixes

### Layer type errors

**Error:** `addIoLayer(): Unknown layer type 'scatter'. Did you mean 'point'?`

You used a type name that doesn't exist. myIO suggests the closest match.
Use `myio_list_chart_types()` to list the supported types.

```{r fix-type, eval = FALSE}
# Wrong
addIoLayer(type = "scatter", ...)

# Right
addIoLayer(type = "point", ...)
```

### Column not found

**Error:** `addIoLayer(): Column 'weight' not found in data. Available columns: mpg, cyl, disp, hp, ...`

The column name in your `mapping` doesn't match any column in `data`.
Check spelling and case — R column names are case-sensitive.

```{r fix-column, eval = FALSE}
# Wrong — 'weight' doesn't exist in mtcars
addIoLayer(mapping = list(x_var = "weight", y_var = "mpg"), ...)

# Right — the column is 'wt'
addIoLayer(mapping = list(x_var = "wt", y_var = "mpg"), ...)

# Check your column names first
names(mtcars)
```

### Invalid transform for type

**Error:** `addIoLayer(): Transform 'lm' is not valid for layer type 'bar'.`

Not all transforms work with all chart types. For example, `lm` only
works with `line` (it produces fitted values for a trend line).

```{r fix-transform, eval = FALSE}
# Wrong — can't apply regression to a bar chart
addIoLayer(type = "bar", transform = "lm", ...)

# Right — use 'line' for trend lines
addIoLayer(type = "line", transform = "lm", ...)
```

### Duplicate layer label

**Error:** `addIoLayer(): Layer label 'data' already exists. Each layer must have a unique label.`

Every layer needs a unique `label`. If you're adding multiple layers,
give each one a distinct name.

```{r fix-label, eval = FALSE}
# Wrong — same label used twice
myIO() |>
  addIoLayer(type = "point", label = "data", ...) |>
  addIoLayer(type = "line", label = "data", ...)

# Right — unique labels
myIO() |>
  addIoLayer(type = "point", label = "scatter", ...) |>
  addIoLayer(type = "line", label = "trend", ...)
```

### Argument type errors

**Error:** `setSlider(): \`param\` must be a single character string, not numeric (123).`

The function tells you exactly what it expected and what you gave it.
Check the argument types in `?setSlider`.

```{r fix-arg, eval = FALSE}
# Wrong — param should be a string, not a number
myIO() |> setSlider(123, "label", 0, 1, 0.5)

# Right
myIO() |> setSlider("ci_level", "Confidence Level", 0.80, 0.99, 0.95)
```

### Missing SharedData

**Error:** `setLinked(): \`shared_data\` must be a SharedData object, not data.frame.`

`setLinked()` requires a `crosstalk::SharedData` object, not a raw
data frame. Wrap your data first.

```{r fix-shared, eval = FALSE}
library(crosstalk)

# Wrong — passing raw data.frame
myIO() |> setLinked(mtcars)

# Right — wrap in SharedData
shared <- SharedData$new(mtcars, key = ~rownames(mtcars))
myIO() |>
  addIoLayer(type = "point", label = "pts",
    data = shared$data(), mapping = list(x_var = "wt", y_var = "mpg")) |>
  setLinked(shared)
```

### Widget type errors

**Error:** `Expected a myIO widget, got 'list'.`

You passed something other than a `myIO()` widget to a setter function.
Make sure the first argument in your pipe is `myIO()`.

```{r fix-widget, eval = FALSE}
# Wrong — list is not a myIO widget
setBrush(list())

# Right — start with myIO()
myIO() |>
  addIoLayer(type = "point", label = "pts",
    data = mtcars, mapping = list(x_var = "wt", y_var = "mpg")) |>
  setBrush()
```

## Chart Renders with Missing Layers

If your chart appears but is missing layers you expected, a JavaScript
validation may have removed them. These errors are not visible in R
by default.

### How to diagnose

1. **Run `myIO_last_error()`** for guidance on where to find the error
2. **Open the browser's developer console** (F12 in most browsers)
3. Look for warnings prefixed with `[myIO]`:
   - `[myIO] Layer 'trend' removed: field 'y_var' must be numeric.`
   - `[myIO] Composition error: Cannot mix standalone chart types with other layers.`

### In Shiny

Read the error reactive input:

```{r shiny-error, eval = FALSE}
observeEvent(input$`myIO-chart-error`, {
  showNotification(input$`myIO-chart-error`, type = "error")
})
```

## Shiny Input Reference

All myIO Shiny inputs follow `myIO-{outputId}-{event}`:

| Input key | When it fires |
|-----------|---------------|
| `myIO-{id}-error` | JS validation or render error |
| `myIO-{id}-rollover` | Hover on data element |
| `myIO-{id}-dragEnd` | Point drag completed |
| `myIO-{id}-brushed` | Brush selection end/clear |
| `myIO-{id}-annotated` | Annotation add/remove/clear |
| `myIO-{id}-slider-{param}` | Slider value change |
